A Large-Scale Study of Polymorphism in R

### 2019-04-05 paper/main.tex
moved poly examples, filled in and edited descriptions

@@

	\item $\square$, \sN: there are a few instances of this ``null'' polymorphism.
	The frequency of these signatures indicates a tendency to create ``optional arguments'', where some argument's default value might be NULL.
	This pattern is quite frequent in R code, \AT{and there are a number of examples sprinkled throughout the paper.}

@@

%
%
%
\section{Examples of Polymorphism}
\label{sec:polyex}

First, Figure~\ref{fig:realex} shows an example of how integers and doubles
are very compatible.  \AT{TODO add matrix (dimensions) to the example.}

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
> 5L + 1L # => 6L, an integer
> 5L + 1.2 # => 6.2, a double (5L coerced to 5.0)
> c(10, 20, 30)[1.2] # => 10, 1.2 coerced to 1L
\end{lstlisting}}\caption{Example of {\it real} type usage.}\label{fig:realex}\end{figure}

Figure~\ref{fig:optnull} shows an example of a function with an optional
argument.  In the function, {\tt x} is a sorted vector of income values, and
{\tt w} is a vector of weights, optionally NULL.  The function computes
fractional ranks required for the computation of some coefficient
implemented by the package.  We see that if {\tt w} is NULL, then default
unit weights are generated.

Figure~\ref{fig:listvec} shows an example of a function with list and vector
polymorphism.  The function takes a list or vector {\tt point} and a data
frame {\tt polyg} representing a polygon.  In it, we see that {\tt point}
can be either a list or a vector, and the function code casts the argument
to a double vector.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
is.point.inside <- function (point, polyg) {
    p <- as.numeric(point) # as.numeric(list(1, 2)) => c(1, 2)
    ...
\end{lstlisting}}\caption{Example of list/vector argument (from {\tt bivrp} package).}\label{fig:listvec}\end{figure}

Figure~\ref{fig:charclos} shows an example of a function with character and
function polymorphism.  \AT{Bluh.}  At a high level, this function
calculates point estimates for an {\tt angmcmc} object (specific to some
packages).  The caller specifies {\tt fn}, either a function or a function
name which will be evaluated on the object samples to estimate parameters.
The call to lookup the function name is on line 10.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
pointest <- function (object, fn = mean, par.name, comp.label, chain.no,  ...) {
    ...
    if (is.character(fn))
        if (fn == "MODE" | fn == "MAP")   do_MAP <- TRUE
        else {
            do_MAP <- FALSE
            fn <- match.fun(fn) # looks up the function name
        }
\end{lstlisting}}\caption{Example of char/closure argument (from {\tt BAMBI} package).}\label{fig:charclos}\end{figure}

\AT{I'll explain this better, maybe?}  In Figure~\ref{fig:dfdbl}, we see a
function that, among other things, takes in some data ({\tt dat}) and a
character string ({\tt spss}) specifying either {\tt "in"} or {\tt "out"}.
Now, depending on the value of {\tt spss}, {\tt dat} will either be a data
frame or a vector of doubles.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
nret.translator <- function(dat, spss="out", ...) {
  ...
  if(identical(spss, "out")) {
    if(!is.vector(dat))  stop(simpleError("...'dat' must be a vector!"))
    ...
  } else {
    ...
    # here, dat is data.frame
    items.idx <- items.idx[order(names(dat[, items.idx]))]
  }
\end{lstlisting}}\caption{Example of a data.frame/(\D) argument (from {\tt klausR} package).}\label{fig:dfdbl}\end{figure}

In Figure~\ref{fig:chardbl}, we see a function which takes in a list ({\tt network}), a vector of indices of that list ({\tt fixIndices}), and a
vector of values ({\tt values}).  The locations specified by {\tt fixIndicies} in {\tt network} are updated with {\tt values}.  Here, {\tt fixIndicies} has been observed to be either a character or double vector:
In R, list indices are typically doubles, but can be characters (if the list
or vector being indexed has a {\tt names} attribute).

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
fixGenes <- function (network, fixIndices, values) {
  ...
  network$fixed[fixIndices] <- as.integer(values)
  ...
\end{lstlisting}}\caption{Example of character/double argument (from {\tt BoolNet} package).}\label{fig:chardbl}\end{figure}%$
In Figure~\ref{fig:matvec}, we see that the function can take in a vector,
but immediately transforms it (transpose, with {\tt t}) into a matrix.  As
an idea, matrix/vector polymorphism seems sensible, as mathematically
vectors are matrices.  R echoes this by making conversion between the two
``types'' easy: \code{as.vector(m)} flattens a matrix \code{m} into a vector
(e.g., \code{as.vector(matrix(2, 2, 2)) == c(2, 2, 2, 2)}), and
\code{as.matrix(v)} builds a {\tt length(v) x 1} matrix (e.g.,
\code{as.matrix(c(1, 2)) == matrix(data=c(1, 2))}).  \AT{TODO fix line break
  in code at end of last sentence}
\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
tee <- function (x, theta, D1, D2, phi)  {
    if (is.vector(x)) x <- t(x)
    ...
\end{lstlisting}}\caption{Example of matrix/(\D) argument (from {\tt calibrator} package).}\label{fig:matvec}\end{figure}
%
%
%
%
\section{Related Work}
\AT{Not sure what the best place for this is, but we should discuss these papers somewhere.}
The idea of measuring polymorphism via some program analysis isn't new.
For instance, some work~\cite{aakerblom2015measuring} ... .
More generally, there is an existing literature on analyzing usage patterns
of language features.  For instance, the dynamic features of Smalltalk were
analyzed in some work~\cite{callau2011howdevelopers}, ...
Other analysis work~\cite{milojkovic2017duck} ... .
\section{Conclusions and Future Work}
\bibliographystyle{boilerplate/ACM-Reference-Format}
\bibliography{bib/biblio,bib/jv,bib/r,bib/new}
\end{document}

>>>>

Before presenting the results of our analysis, we will quickly survey examples of real, polymorphic R functions that we have analyzed, to build up a global sense of how types manifest themselves in current R code.

%
%
%
%
%
%
\section{Examples of Polymorphism}
\label{sec:polyex}

In this section, we will present an eclectic mix of R functions which exhibit polymorphic behavior at {\bf L0}.

We begin by referring the reader to Figure~\ref{fig:realex}, which showcases the compatibility of integers and doubles.
Coercion from integer to double is performed automatically where appropriate (e.g., here when a double and integer are added together), and likewise coercion from double to integer is automatically performed when integers are strictly required (e.g., here when doubles are being used to index a vector or list).

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
> 5L + 1L # => 6L, an integer
> 5L + 1.2 # => 6.2, a double (5L coerced to 5.0)
> c(10, 20, 30)[1.2] # => 10, 1.2 coerced to 1L
\end{lstlisting}}\caption{Example of \sD, \sI type usage.}\label{fig:realex}\end{figure}

Figure~\ref{fig:optnull} shows an example of a function with an optional NULL argument. 
In the function, {\tt x} is a sorted vector of income values, and {\tt w} is a vector of weights with default value NULL.
The function computes fractional ranks required in other package functions.
We see that if {\tt w} is NULL, then default unit weights are generated.
This pattern of default NULL (or NA) arguments is very prevalent in R.

Figure~\ref{fig:listvec} shows an example of a function with list and vector
polymorphism.  The function takes a list or vector {\tt point} and a data
frame {\tt polyg} representing a polygon.  In it, we see that {\tt point} is explicitly converted
to a vector (via the \code{as.numeric} function) regardless of its type.
This showcases the ease with which lists and vectors may be converted in R (the corresponding vector-to-list function is \code{as.list}).

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
is.point.inside <- function (point, polyg) {
    p <- as.numeric(point) # as.numeric(list(1, 2)) => c(1, 2)
    ...
\end{lstlisting}}\caption{Example of \l/\D argument (from {\tt bivrp} package).}\label{fig:listvec}\end{figure}

Figure~\ref{fig:charclos} shows an example of a function with character and
function polymorphism.  At a high level, this function
calculates point estimates for an {\tt angmcmc} object (specific to some
packages).  The caller specifies {\tt fn}, which is either a function {\it or} the name of a function which will be evaluated on the object samples to estimate parameters.
If {\tt fn} is a string, the function lookup is performed by \code{match.fun}.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
pointest <- function (object, fn = mean, par.name, comp.label, chain.no,  ...) {
    ...
    if (is.character(fn))
        if (fn == "MODE" | fn == "MAP")   do_MAP <- TRUE
        else {
            do_MAP <- FALSE
            fn <- match.fun(fn) # looks up the function name
        }
\end{lstlisting}}\caption{Example of \sC/\sF argument (from {\tt BAMBI} package).}\label{fig:charclos}\end{figure}

In Figure~\ref{fig:dfdbl}, we see a function that, among other things, takes in some data ({\tt dat}) and a
character string ({\tt spss}) specifying either {\tt "in"} or {\tt "out"}.
Here, the type of {\tt dat} is {\it dependent} on the value of {\tt spss}:
if {\tt spss == "in"}, {\tt dat} must be a data frame, and otherwise if {\tt spss == "out"}, {\tt dat} must be a vector.
Our analysis does not capture this granularity of dependency, as we only collect type information.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
nret.translator <- function(dat, spss="out", ...) {
  ...
  if(identical(spss, "out")) {
    if(!is.vector(dat))  stop(simpleError("...'dat' must be a vector!"))
    ...
  } else {
    ...
    # here, dat is data.frame
    items.idx <- items.idx[order(names(dat[, items.idx]))]
  }
\end{lstlisting}}\caption{Example of a \df/\D argument (from {\tt klausR} package).}\label{fig:dfdbl}\end{figure}

In Figure~\ref{fig:chardbl}, we see a function which takes in a list ({\tt network}), a vector of indices of that list ({\tt fixIndices}), and a
vector of values ({\tt values}).  The locations specified by {\tt fixIndicies} in {\tt network} are updated with {\tt values}.  Here, {\tt fixIndicies} has been observed to be either a character or double vector:
In R, list indices are typically doubles, but can be characters (if the list
or vector being indexed has a {\tt names} attribute).
Note that type level {\bf L2} can capture the presence of a {\tt names} attribute.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
fixGenes <- function (network, fixIndices, values) {
  ...
  network$fixed[fixIndices] <- as.integer(values)
  ...
\end{lstlisting}}\caption{Example of \C/\D argument (from {\tt BoolNet} package).}\label{fig:chardbl}\end{figure}

In Figure~\ref{fig:matvec}, we see that a function which can take in a vector,
but immediately transforms it (transpose, with {\tt t}) into a matrix.  As
an idea, matrix/vector polymorphism seems sensible, as mathematically
vectors {\it are} one-dimensional matrices.  R echoes this by making conversion between the two
types easy: \code{as.vector(m)} flattens a matrix \code{m} into a vector
(e.g., \code{as.vector(matrix(2, 2, 2)) == c(2, 2, 2, 2)}), and
\code{as.matrix(v)} builds a {\tt length(v) x 1} matrix (e.g.,
\code{as.matrix(c(1, 2)) == matrix(data=c(1, 2))}).  \AT{TODO fix line break
  in code at end of last sentence}

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
tee <- function (x, theta, D1, D2, phi)  {
    if (is.vector(x)) x <- t(x)
    ...
\end{lstlisting}}\caption{Example of matrix/(\D) argument (from {\tt calibrator} package).}\label{fig:matvec}\end{figure}

Throughout this short section, we have shown real examples of polymorphic R functions.
Our goal is to capture these instances of polymorphism across a large subset of existing R code, and to highlight and discuss patterns which arise.
We present the results of our analysis in the following section.

@@

	\item $\square$, \sN: there are a few instances of this ``null'' polymorphism.
	The frequency of these signatures indicates a tendency to create ``optional arguments'', where some argument's default value might be NULL.
	This pattern is quite frequent in R code, and an example can be found in Figure~\ref{fig:optnull}.

@@

%
%
%
%
\section{Related Work}
\AT{Not sure what the best place for this is, but we should discuss these papers somewhere.}
The idea of measuring polymorphism via some program analysis isn't new.
For instance, some work~\cite{aakerblom2015measuring} ... .
More generally, there is an existing literature on analyzing usage patterns
of language features.  For instance, the dynamic features of Smalltalk were
analyzed in some work~\cite{callau2011howdevelopers}, ...
Other analysis work~\cite{milojkovic2017duck} ... .
\section{Conclusions and Future Work}
\bibliographystyle{boilerplate/ACM-Reference-Format}
\bibliography{bib/biblio,bib/jv,bib/r,bib/new}
\end{document}
